---
title: "Bap1 fgsea Pathway Analysis"
subtitle: "Glia DEGs KO vs. WT, across all clusters"
author: "Kat Beigel (beigelk@chop.edu)"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# utility
library(tidyverse)
library(readr)
library(xlsx)
library(tools)
library(gtools)
library(ggplot2)
library(scales)
library(mgsub)
library(msigdbr)
library(viridis)
library(DT)
library(data.table)

# analysis
library(fgsea)

```

# Project set up

```{r Directories and project info}

# Main dir for the project
base_dir = "DEGPathwayAnalysis/"

# Where to load DEG lists from
data_dir = "Seurat/DifferentialGeneExpression/"

# Where to write results
res_dir = "DEGPathwayAnalysis/fgsea/"

infile = "Glia_GeneExpression_ConditionsCompared_V1.xlsx"

version = "V1"
proj = "Bap1_Glia_KO_v_WT"
analysis_name = "fgsea"
name = "all clusters"

```

``` {r Load data}

#  data_files = list.files(data_dir)

# Load in significant genes from DEG analysis in Seurat
deg_tbl = read.xlsx(paste0(data_dir, infile), sheetIndex = 1) %>%
  # Fix col names
  dplyr::rename(symbol = NA.) %>%
  # Drop unneeded columns
  dplyr::select(-c(NA..1)) %>%
  mutate(symbol = gsub('\\.1', '', symbol)) %>%
  # This data sheet reports all positive fold changes,
  # need to convert the "positive in WT" to negative
  mutate(
    avg_logFC = case_when(
      cluster == "WT" ~ -(avg_logFC),
      .default = avg_logFC)
  ) %>%
  # Update the cluster to say KO
  mutate(
    cluster = "KO"
  ) %>%
  # Need to change the logFC to log2FC as that is more standard
  mutate(
    # https://www.statology.org/antilog-in-r/
    avg_fold = exp(avg_logFC),
    avg_log2FC = log2(avg_fold)
  ) %>%
  dplyr::arrange(desc(avg_logFC))

```

# DEG table 

Data in table has been modified to show positive/negative values for KO (KO v. WT condition) by flipping the sign for logFC if the cluster column == WT, as the original file had only positive logFC with the directionality of the change indicated by the cluster identity.

Table has been edited to also show fold change and log2FC calculated from logFC (natural log) values.



```{r Input table}

datatable(deg_tbl %>%
            tibble() # %>%
            # mutate(across(where(is.numeric), \(x) round(x, 3))
            )

fwrite(deg_tbl, file = paste0(res_dir, "csvs/", "DEGs", "_", proj, "_", gsub(' ', '_', name), "_", version, ".csv"),
       sep = ",", sep2 = c("", "|", ""))

```

``` {r Functions}

# Imports MSigDB datasets
ImportGenesetData <- function(species_name, collection, subcollection){
  if (missing(subcollection)) genesets <- msigdbr(species = species_name, category = collection)
  else genesets <- msigdbr(species = species_name, category = collection, subcategory = subcollection)
  genesets_list <- split(x = genesets$gene_symbol, f = genesets$gs_name)
  genesets_info <- split(x = genesets$gs_description, f = genesets$gs_name)
  for (i in 1:length(genesets_list)){
    genesets_list[[i]] <- unique(genesets_list[[i]])
    genesets_info[[i]] <- unique(genesets_info[[i]])
  }
  list(genesets_list, genesets_info)
}

```

# Pathways to analyze from MSigDB (via msigdbr)

``` {r Load gene sets from msigdbr}
# Load data from msigdb via msigdbr
# https://igordot.github.io/msigdbr/

species = "Mus musculus"

hallmark <- ImportGenesetData(species, "H")

c2cp <- c(ImportGenesetData(species, "C2", "CP:BIOCARTA"),
          ImportGenesetData(species, "C2", "CP:KEGG"),
          ImportGenesetData(species, "C2", "CP:PID"),
          ImportGenesetData(species, "C2", "CP:REACTOME"),
          ImportGenesetData(species, "C2", "CP:WIKIPATHWAYS"))

pathway_genesets_list <- list(hallmark, c2cp)

pathway_genesets_names <- c("Hallmark", "C2.CP")

```

```{r Preparing pathway analysis}


complete_list_genesets = c(pathway_genesets_list[[1]][[1]],
                           pathway_genesets_list[[2]][[1]],
                           pathway_genesets_list[[2]][[3]],
                           pathway_genesets_list[[2]][[5]],
                           pathway_genesets_list[[2]][[7]],
                           pathway_genesets_list[[2]][[9]])

```

# fgsea Analysis

```{r Run fgsea, fig.width=15, fig.height=10}

deg_tbl_ordered = deg_tbl %>%
  arrange(desc(avg_log2FC))

print(paste0("Analysis: ", name))

# Make gene list
gene_list = deg_tbl_ordered %>% pull(avg_log2FC)
names(gene_list) <- deg_tbl_ordered %>% pull(gene)
barplot(gene_list, cex.lab = 1, cex.names = 0.1, las = 2, ylim = c(-10, 10))

# fgsea should be used with set.seed() for reproducibility
# https://github.com/ctlab/fgsea/issues/12
set.seed(2345678)
fgsea_res <- fgsea(pathways = complete_list_genesets,
                   stats = gene_list,
                   minSize = 3,
                   maxSize = 500,
                   eps = 0.0)

fgsea_res_filt = fgsea_res %>%
  filter(padj < 0.05) %>%
  mutate(database = gsub("_(.*?)$", '', .$pathway)) %>%
  mutate(shortname = gsub("^(.*?)_", '', .$pathway)) %>%
  mutate(prettyname = gsub('_', ' ', .$shortname))

```

# fgsea Results

## Figure of fgsea results: Significant Pathways

```{r figure, fig.width=15, fig.height=15}

if (dim(fgsea_res_filt)[1] != 0){
  
  lolli = ggplot(
    fgsea_res_filt,
    aes(x = reorder(prettyname, -NES), y = NES)) +
    # add segments based on NES and color based on padj
    geom_segment(aes(x = reorder(prettyname, -NES),
                     xend = reorder(prettyname, -NES),
                     y = 0,
                     yend = NES,
                     color = padj)) +
    # adds points with specified colors and sizes to the end of the segments based on values in the dataframe
    geom_point(aes(color = padj,
                   size = size)) +
    # manually set the color and scale of p.adjust if needed
    scale_color_viridis(limits = c(0, 0.05),
                        direction = 1,
                        option = "turbo",
                        end = 0.9, begin = 0.25) +
    # flip the x and y coords to put pathways on the y-axis if you'd like
    coord_flip() +
    # # changing theme to light
    theme_light() +
    # so that they're in alphabetical order and wrap the text after 30 characters to fit better.
    scale_x_discrete(label = function(x) str_wrap(x, width = 60)) +
    # split into panels based on conditions/comparisons (ontology in this case) and put panel label on right
    facet_grid(database ~ ., scales = "free_y", space = "free_y") +
    # if needed, set y-limits for NES
    # ylim(-4.5, 4.5) +
    # add vertical line at 0 NES
    geom_hline(yintercept = 0, color = "black") +
    # changing theme elements
    theme(
      #panel.grid.major.y = element_blank(),
      panel.border = element_rect(colour = "black"),
      axis.ticks = element_line(color = "black"),
      axis.title.y = element_blank(),
      #legend.title.align = 0.5,
      legend.key.height = unit(1, 'cm')
    ) + 
    # change theme elements of panel titles (can change colors and fonts)
    theme(strip.background = element_rect(fill = "grey90", color = "black"),
          strip.text = element_text(colour = 'black', face = "bold"),
          plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5, size = 8)) +
    # change legend labels and y-axis label
    labs(title = paste0("fgsea Significant Pathways for DEGs in ",
                        gsub('_', ' ', gsub("^(.*?)_", '', name))),
         color = "Adjusted p-value",
         size = "Number of genes in pathway",
         y = "Normalized Enrichment Score")
  
  print(paste0("Results: significant pathways for ", name))
  print(lolli)
  
  pdf(file = paste0(res_dir, analysis_name, "_", proj, "_", name, "_", version, ".pdf"),
      height = 12, width = 10)
  print(lolli)
  dev.off()
  
  png(file = paste0(res_dir, analysis_name, "_", proj, "_", name, "_", version, ".png"),
      height = 12, width = 10, units = "in", res = 300)
  print(lolli)
  dev.off()
  
} else {
  print(paste0("Results: No significant pathways for ", name))
}

```

## Table of fgsea results: Significant Pathways

```{r Table of results}

fgsea_res_filt %>% datatable()

fwrite(fgsea_res_filt %>% dplyr::select(-prettyname),
       file = paste0(res_dir, "csvs/", analysis_name, "_", proj, "_", gsub(' ', '_', name), "_", version, ".csv"),
       sep = ",", sep2 = c("", "|", ""))

```

# Reference for gene sets tested

```{r List of gene sets tested}
complete_list_genesets
```

```{r Session info}

sessionInfo()

```
